---
title: "Format decimals to match"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Format decimals to match}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(
  datatable.print.nrows = 15,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

If we use `format_power()` to create powers-of-ten notation in a table, the font face and size of the inline math output is usually different from that of unformatted numbers. 

The purpose of `format_decimal()` is to delimit such numbers to match the powers-of-ten numbers.  








## Practice data table

Using the `atmos` data set include with *formatdown* with various atmospheric properties as a function of height above sea level. 

```{r}
library("formatdown")
library("data.table")

DT <- copy(atmos)
knitr::kable(DT, align = "r")
```




## Power-of-ten column 

Of these variables, density best lends itself to powers-of-ten notation. For this range of values, I select a fixed exponent of $-4$. 

```{r}
# Apply power of ten notation 
DT$dens <- format_power(DT$dens, digits = 3, set_power = -4)

# Render in document
knitr::kable(DT, align = "r")
```

The other numerical columns are OK except the font face and size doesn't match the density column. In this example, I the remaining columns one at a time. 






## Integer column

Changing the units of altitude to km, ensure zero decimal places. 

```{r}
DT$alt <- format_decimal(DT$alt / 1000, 0)

# Render in document
knitr::kable(DT, align = "r")
```






## Integer with big mark column

The pressure column spans several orders of magnitude. I'll try formatting all but the last two rows as integers with a "big mark"  and the last two rows with two decimal places. 

```{r}
# Create a temporary formatted column
DT[pres <= 10,              temp_pres := format_decimal(pres, 2)]
DT[pres > 10 & pres <= 100, temp_pres := format_decimal(pres, 1)]
DT[pres > 100,              temp_pres := format_decimal(pres, 0, big_mark = ",")]

# Overwrite the original and delete the temporary column
DT[, pres := temp_pres]
DT[, temp_pres := NULL]

# Render in document
knitr::kable(DT, align = "r")
```






## Decimal columns

The temperature and speed of sound columns can both be formatted to one decimal place. 
```{r}
cols_to_format <- c("temp", "sound")
DT[, (cols_to_format) := lapply(.SD, function(x) format_decimal(x, 1)), .SDcols = cols_to_format]

# Render in document
knitr::kable(DT, align = "r")
```






## `knitr::kable()` options

I use some `knitr::kable()` arguments to change the header and add a caption. 


```{r}
# Render in document
knitr::kable(DT, 
             align = "r", 
             caption = "Table 1. Atmospheric varables as a function of altitude", 
             col.names = c("Altitude [km]", 
                           "Temperature [K]", 
                           "Pressure [Pa]", 
                           "Density [kg/m$^3$]", 
                           "Speed of sound [m/s]"))
```







