---
title: "Format powers of ten"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
link-citations: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Format powers of ten}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("formatdown")
library("data.table")

# data.table printout
options(
  datatable.print.nrows = 15,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

The notation we use to represent large and small numbers depends on the communications context. In a computer script, for example, we might encode Avogadro's number $(N_A)$ as `6.0221*10^23`. A computer printout of this number would typically use E-notation, as in `6.0221E+23`.

In professional technical communications, however, computer syntax should be avoided---the asterisk (`*`) and carat (`^`) in `6.0221*10^23` encode computer *instructions*, not *mathematics.* And while E-notation (`6.0221E+23`) has currency in some discourse communities, the general convention in technical communications is to format large and small numbers using powers-of-ten notation of the form, 

$$
a \times 10^n, 
$$

The $\times$ multiplication symbol is a convention in powers-of-ten notation even though it is usually avoided in other contexts. Two forms of this notation are in common use [@Chase:2021, 63--67]: 

- *Scientific notation* &nbsp; where $n$ is an integer and $a$ is any real number such that $1\leq{a}<10$, for example, $N_A =$ `r format_power(6.0221E+23, 5, format = "sci")`

- *Engineering notation* &nbsp; where $n$ is a multiple of 3 and $a$ is any real number such that $1\leq{a}<1000$, as in $N_A =$ `r format_power(6.0221E+23, 5)`

When exponents are in the neighborhood of zero, for example, $n \in \{-1, 0, 1, 2\}$, decimal notation may be preferred.  Thus decimal values 0.1234, 1.234, 12.34, and 123.4 will often be printed as-is. The limits of this set can be adjusted to suit one's audience and rhetorical goals. 

## Markup

In R Markdown [@rmd:2022] and Quarto Markdown [@qmd:2022], we use an inline equation markup delimited by `$...$` to create a math expression in the output document. For example, the markup for Avogadro's number is given by, 

        $N_A = 6.0221\times{10}^{23}$

which yields $N_A = 6.0221 \times 10^{23}$ in the document. To *program* the markup, however, we enclose the markup as a character string, that is, 

        "$N_A = 6.0221\\times{10}^{23}$"

which requires the backward slash `\` to be "escaped", hence the product symbol is denoted `\\times`. This is the form of the output produced by `format_power()`. 






## `format_power()`

Given a number, a numerical vector, or a numerical column from a data frame, `format_power()` converts the numbers to character strings of the form, 

        "$a\\times{10}^{n}$"
 
where `a` is the coefficient and `n` is the exponent. The user can specify the number of significant digits, scientific or engineering format, and the range over which decimal notation is enforced. 
 
*Arguments.*

- **x** &nbsp; Numerical vector to be formatted. Can be a scalar, a vector, or a column from a data frame. 

- **digits** &nbsp; Positive nonzero integer to specify the number of significant digits. Default is 3. 

- **format** &nbsp; Possible values are "engr" (engineering notation) and "sci" (scientific notation). Default is "engr". 

- **limits** &nbsp; Numeric vector of length two, default `c(0.1, 1000)`,  specifying the range $p\leq{a}<q$ over which number $a$ is printed without power-or-ten notation. 


If you are writing your own script to follow along, we use these packages in this vignette:

```{r}
library("formatdown")
library("data.table")
```

*Equivalent usage.*  &nbsp;  The first two arguments do not have to be named if the argument order is maintained. 

```{r}
# Arguments named
(x <- format_power(x = 6.0221E+23, digits = 3))

# Arguments unnamed
y <- format_power(6.0221E+23, 3)

# Implicit use of default argument 
z <- format_power(6.0221E+23)

# Demonstrate equivalence
all.equal(x, y)
all.equal(x, z)
```






## Examples

*Scalar.* &nbsp;  Use with inline R code.  

```{r}
format_power(101100)
```

which, in an Rmd or qmd document, is rendered as `r format_power(101100)`. 

<br>*Vector.* &nbsp;  

```{r}
x <- c(1.2222e-6, 2.3333e-5, 3.4444e-4, 4.1111e-3, 5.2222e-2, 6.3333e-1,
       7.4444e+0, 8.1111e+1, 9.2222e+2, 1.3333e+3, 2.4444e+4, 3.1111e+5, 4.2222e+6)
format_power(x)
```

is rendered as `r format_power(x)[1:3]`, etc.  

<br>*Significant digits.* &nbsp; Argument does not have to be named. 

```{r}
format_power(x[1], 3)
format_power(x[1], 4)
```

are rendered as 

- `r format_power(x[1], 3)`
- `r format_power(x[1], 4)`

<br>*Format.* &nbsp; Argument, if included, must be named. 

```{r}
format_power(x[3], format = "sci")
format_power(x[3], format = "engr")
```

are rendered as 

- `r format_power(x[3], format = "sci")`
- `r format_power(x[3], format = "engr")`


<br>*Limits.* &nbsp; Argument, if included, must be named.   

```{r}
format_power(x[6], limits = c(0.1, 1000))
format_power(x[6], limits = c(1, 1000))
```

are rendered as 

- `r format_power(x[6], limits = c(0.1, 1000))`
- `r format_power(x[6], limits = c(1, 1000))`

To compare the effects across many orders of magnitude, we treat the example vector but convert it to a one-column data frame for display, rendered using using `knitr::kable()`, 

```{r}
# Format using defaults and convert to a data frame
DT <- data.table(scientific = format_power(x, format = "sci"), 
                 engineering = format_power(x))
knitr::kable(DT, align = "rr")
```


<br>*Data frame.* &nbsp; Using `density`, a data frame included with formatdown that contains columns of class Date, character, factor, numeric, and integer.

```{r}
density
```

which is rendered un-formatted as

```{r}
#| echo: false
knitr::kable(density, align = "ccrrr")
```

Treating a column as a vector, 

```{r}
# Copy to avoid "by reference" changes to density  
DT <- copy(density)

# Format as a vector
format_power(DT$p_Pa, digits = 4)
```

is rendered as `r format_power(DT$p_Pa, digits = 4)`. 

Treating a column within a data frame, 

```{r}
# Copy to avoid "by reference" changes to density  
DT <- copy(density)

# Format one column, retain all columns
DT$p_Pa <- format_power(DT$p_Pa, digits = 4)
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "r")
```

Using `lapply()` to select and treat multiple columns from a data frame, 

```{r}
# Copy to avoid "by reference" changes to density 
DT <- copy(density)

# Identify columns to format
cols_we_want <- c("T_K", "p_Pa", "density")

# Select and format.  
DT <- DT[, lapply(.SD, function(x) format_power(x, 4)), .SDcols = cols_we_want]
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "r")
```

Repeat, but retain all columns,  

```{r}
# Copy to avoid "by reference" changes to density 
DT <- copy(density)

# Identify columns to format
cols_we_want <- c("T_K", "p_Pa", "density")

# Format selected columns, retain all columns
DT <- DT[, (cols_we_want) := lapply(.SD, function(x) format_power(x, 4)), .SDcols = cols_we_want]
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "ccrrr")
```

If different columns should be reported with different significant digits, treat the columns separately. 

```{r}
# Copy to avoid "by reference" changes to density 
DT <- copy(density)

# Format selected columns, retain all columns with signif digits = 5
cols_we_want <- c("T_K", "density")
DT <- DT[, (cols_we_want) := lapply(.SD, function(x) format_power(x, 5)), .SDcols = cols_we_want]

# Individually format one column with signif digits = 4
DT$p_Pa <- format_power(DT$p_Pa, 4)
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "ccrrr")
```

*Non-numeric input.* &nbsp; Applying the function to non-numeric objects produces errors. For example, "Date" class. 

```{r}
#| error: true
x <- density$date
format_power(x)
```

Error on "character" class, 

```{r}
#| error: true
x <- density$trial
format_power(x)
```

Error on "factor" class, 

```{r}
#| error: true
x <- density$humidity
format_power(x)
```




## References



