---
title: "Format numbers"
vignette: >
  %\VignetteIndexEntry{Format numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/information-science-and-technology.csl
link-citations: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# to knit "child" Rmd files
knitr::opts_knit$set(root.dir = "../")

library(formatdown)
library(data.table)
library(knitr)
library(stringr)

options(
  datatable.print.nrows = 15,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```


```{r}
#| echo: false

formatdown_options(size = "small")

prefix_1 <- c("peta", "tera", "giga", "mega", "kilo")

symbol_1 <- c("P", "T", "G", "M", "k")
symbol_1 <- format_text(symbol_1, face = "italic")

x <- 10^seq(from = 15, to = 3, by = -3)
value_1 <- format_numbers(x, digits = 1, format = "engr", omit_power = NULL)
value_1 <- sub("1 \\\\times ", "", value_1)

prefix_2 <- c("milli", "micro", "nano", "pico", "femto")

symbol_2 <- c("m", "\\mu", "n", "p", "f")
symbol_2 <- format_text(symbol_2, face = "italic")

x <- 10^seq(from = -3, to = -15, by = -3)
value_2 <- format_numbers(x, 1, omit_power = NULL)
value_2 <- sub("1 \\\\times ", "", value_2)

DT <- data.table(prefix_1, symbol_1, value_1, prefix_2, symbol_2, value_2)
knitr::kable(DT,
  align = "lcllcl",
  col.names = rep(c("Prefix", "Symbol", "Value"), 2)
)
```

The rationale for the formatdown package is formatting numbers in power-of-ten notation in inline R code or tabulated columns of data frames. Other features of the package provide tools for typesetting non-power-of-ten columns to match. In this vignette, we discuss the primary formatting function `format_numbers()` and its convenience wrappers for scientific, engineering, and decimal notation.

## Types of notation

Notation to represent large and small numbers depends on the mode of communication. In a computer script, for example, we might encode the Avogadro constant as `N_A = 6.0221*10^23`. The asterisk (\*) and caret (\^) in this expression, however, communicate *instructions to a computer*, not *syntactical mathematics*. And while scientific E-notation (`6.0221E+23`) has currency in some discourse communities, power-of-ten notation, e.g., $\small N_A = 6.0221 \times 10^{23}$, is the conventional format for professional technical communication. 

Power-of-ten notation is expressed,   

$$
\small a \times 10^n, 
$$

where $\small a$ is the coefficient in decimal form and the exponent $\small n$ is an integer. Two formats are in common use [@Chase:2021, 63--67]: 

- *scientific:* &nbsp; $\small 1\leq{|a|} < 10$, e.g., $\small N_A =$ `r format_sci(6.0221E+23, 5)`. 

- *engineering:* &nbsp; $\small  1\leq{|a|} < 1000$ and $\small n$ is a multiple of 3, e.g., $\small N_A =$ `r format_engr(6.0221E+23, 5)`. 

The utility of the engineering form follows from the SI prefixes for physical units such as "mega-", "kilo-", "milli-", etc., corresponding to powers of 10 that are integer multiples of three. 

<br>

*Notes on syntax.* &nbsp; Programming symbols are not necessarily mathematical symbols:

- asterisk (\*). Programming symbol for multiplication, e.g., `x = a * b` or `y = a * (b + c)`. In the grammar of mathematics, multiplication is indicated by the symbol ($\times$) when needed.   

- caret (\^). Programming symbol for exponentiation, e.g., `x = y^2` or `z = 10^-3`. In the grammar of mathematics, exponents are typeset as superscripts, e.g., $\small x = y^2$ or $\small z = 10^{-3}$.

- multiplication ($\small\times$). Mathematical symbol for multiplication, not to be confused with the letter "x". Generally omitted when the meaning is clear, e.g., $\small x=ab$ or $\small y=a(b+c)$, but conventionally included in power-of-ten notation, e.g., $\small 6.0221 \times 10^{23}$. When a comma is used as the decimal marker, multiplication may be indicated by the half-high dot, e.g., $\small 6,0221 \cdot 10^{23}$.

<br>

*Decimal subsets.* &nbsp; In a vector of numbers formatted in power-of-ten form, the decimal form may be preferred for any subset of values with exponents near zero, e.g., $\small n \in \{-1, 0, 1, 2\}$. 


```{r}
#| echo: false

x <- 3.12 * 10^seq(-3, 4)

sci <- format_sci(x, 3, omit_power = NA)
sci_omit <- format_sci(x, 3)
DT <- data.table(sci, sci_omit)

knitr::kable(DT,
  align = "r",
  caption = "Decimal form may be preferred for a subset",
  col.names = c(
    "scientific notation",
    "subset in decimal form"
  )
)
```

<br>

*Decimal columns.* &nbsp; A table of numeric information can include columns formatted in both power-of-ten notation and decimal notation. For example, a  table of atmospheric properties shown below has altitude in integer form, temperature in decimal form,  and density in power-of-ten engineering notation (except for those values with exponents near zero). 

```{r}
#| echo: false

DT <- atmos[alt < 150, .(alt, temp, dens)]

DT$alt <- format_dcml(DT$alt, 2)
DT$temp <- format_dcml(DT$temp, 5)
DT$dens <- format_engr(DT$dens, 3)
knitr::kable(DT,
  align = "r",
  caption = "Properties of the atmosphere",
  col.names = c(
    "Altitude (km)",
    "Temperature (K)",
    "Density (kg/m$^3$)"
  )
)

formatdown_options(reset = TRUE)
```

The purpose of the decimal format in formatdown is to match the font face and size of decimal columns to those of the power-of-ten columns. If no power-of-ten columns are used, of course, decimal columns can be displayed as-is or formatted using other R tools. 

<br>

*Packages.* &nbsp; If you are writing your own script to follow along, we use the following packages in this vignette. Data frame operations are performed with data.table syntax. Some users may wish to translate the examples to use  base R or dplyr syntax.  

```{r}
#| echo: true
#| eval: false

library("formatdown")
library("data.table")
library("knitr")
```

## Markup {#markup}

We format numbers as inline math expressions  delimited by `$ ... $` or the optional `\( ... \)`. For example, the Avogadro constant is marked up as

        $6.0221 \times 10^{23}$, 

where the `\times` macro creates the multiplication symbol ($\small\times$). This math markup, as an inline equation in an R markdown document, renders as: $\small 6.0221 \times 10^{23}$. To *program* the markup, however, we enclose it in quote marks as a character string, that is, 

        "$6.0221 \\times 10^{23}$", 

which requires us to "escape" the backslash in `\times` by adding an extra backslash. When the optional font size argument is assigned, formatdown adds a LaTeX-style sizing macro such as `\small` or `\large`, for example, 

        "$\\small 6.0221 \\times 10^{23}$", 

where again the markup includes an extra backslash.  



## `format_sci()` {#format_sci}

Converts numbers to character strings in power-of-ten form, 

        "$a \\times 10^{n}$" 

where $\small a$ is the coefficient and $\small n$ is the exponent. `format_sci()` is a wrapper for the more general function `format_numbers()`. For a subset of values with exponents near zero, e.g., $\small n \in \{-1, 0, 1, 2\}$, the output is in decimal form, 

        "$a$"

<br>

*Usage.* &nbsp; 

```r
format_sci(x,
           digits = 4,
           ...,
           omit_power = c(-1, 2),
           set_power = NULL,
           delim          = formatdown_options("delim"),
           size           = formatdown_options("size"),
           decimal_mark   = formatdown_options("decimal_mark"),
           small_mark     = formatdown_options("small_mark"),
           small_interval = formatdown_options("small_interval"), 
           whitespace     = formatdown_options("whitespace"), 
           multiply_mark  = formatdown_options("multiply_mark"))
```

- Arguments before the dots do not have to be named if argument order is maintained. 
- Arguments after the dots (`...`) must be named. 
- The arguments assigned via `formatdown_options()` can be reset by the user  locally in a function call or globally using `formatdown_options()`. 
 

<br>

*Examples.*  &nbsp; These early examples are shown with default arguments. Arguments are explored more fully starting with [Numeric input](#numeric-input) section. 

```{r}
# 1. Avogadro constant
L <- 6.0221E+23
format_sci(L)

# 2. Elementary charge
e <- 1.602176634e-19
format_sci(e)
```

Examples 1 and 2 (in inline code chunks) render as,    

1. The Avogadro constant is $\small L =$ `r format_sci(L, size = "small")` $\small \mathit{mol}^{-1}$.
1. The elementary charge constant is $\small e =$ `r format_sci(e, size = "small")` $\small C$. 




## `format_engr()` {#format_engr}

Similar to `format_sci()` except using engineering notation, i.e., exponents are multiples of 3. 

<br>

*Usage.* 

```r
format_engr(x,
            digits = 4,
            ...,
            omit_power = c(-1, 2),
            set_power = NULL,
            delim          = formatdown_options("delim"),
            size           = formatdown_options("size"),
            decimal_mark   = formatdown_options("decimal_mark"),
            small_mark     = formatdown_options("small_mark"),
            small_interval = formatdown_options("small_interval"), 
            whitespace     = formatdown_options("whitespace"), 
            multiply_mark  = formatdown_options("multiply_mark"))
```

<br>

*Examples.*  &nbsp; (with default arguments)

```{r}
# 3. Avogadro constant
format_engr(L)

# 4. Elementary charge
format_engr(e)
```

Examples 3 and 4 render as,   

3. The Avogadro constant is $\small L =$ `r format_engr(L, size = "small")` $\small \mathit{mol}^{-1}$.
3. The elementary charge constant is $\small e =$ `r format_engr(e, size = "small")` $\small C$. 




## `format_dcml()` {#format_dcml}

A wrapper for the more general function `format_numbers()`; converts numbers to  character strings in decimal form, 

        "$a$"

 where $\small a$ is the decimal value.  

<br>

*Usage.* 

```r
format_dcml(x,
            digits = 4,
            ...,
            size           = formatdown_options("size"),
            delim          = formatdown_options("delim"),
            decimal_mark   = formatdown_options("decimal_mark"),
            big_mark       = formatdown_options("big_mark"),
            big_interval   = formatdown_options("big_interval"),
            small_mark     = formatdown_options("small_mark"),
            small_interval = formatdown_options("small_interval"), 
            whitespace     = formatdown_options("whitespace"))
```

<br>

*Examples.*  &nbsp; (with default arguments)

```{r}
# 5. Speed of light in a vacuum
c <- 299792458
format_dcml(c)

# 6. Molar gas constant
R <- 8.31446261815324
format_dcml(R)
```

Examples 5 and 6 render as,  

5. The speed of light in a vacuum is $\small c =$ `r format_dcml(c, size = "small")` $\small\mathit{m/s}$.
5. The molar gas constant is $\small R =$ `r format_dcml(R, size = "small")` $\small\mathit{J}\cdot\mathit{K}^{-1}\mathit{mol}^{-1}$. 





## `format_numbers()` {#format_numbers}

`format_numbers()` is the general-purpose formatting function called by `format_sci()`, `format_engr()`, and `format_dcml()`. The general function can be used instead of the convenience functions simply by setting its `format` argument to `"sci"`, `"engr"` (default), or `"dcml"`. 


<br>

*Usage.* 

```r
format_numbers(x,
               digits = 4,
               format = "engr",
               ...,
               omit_power = c(-1, 2),
               set_power = NULL,
               delim          = formatdown_options("delim"),
               size           = formatdown_options("size"),
               decimal_mark   = formatdown_options("decimal_mark"),
               big_mark       = formatdown_options("big_mark"),
               small_mark     = formatdown_options("small_mark"),
               big_interval   = formatdown_options("big_interval"),
               small_interval = formatdown_options("small_interval"), 
               whitespace     = formatdown_options("whitespace"), 
               multiply_mark  = formatdown_options("multiply_mark"))
```


*Examples.*  &nbsp; Reproducing some of the earlier examples using `format_numbers()`.

```{r}
# 7. Scientific
format_numbers(L, format = "sci")

# 8. Engineering
format_numbers(e, format = "engr")

# 9. Decimal
format_numbers(R, format = "dcml")
```

Examples 7--9 render as,  

7. The Avogadro constant is $\small L =$ `r format_numbers(L, format = "sci", size = "small")` $\small \mathit{mol}^{-1}$. 
7. The elementary charge constant is $\small e =$ `r format_numbers(e, format = "engr", size = "small")` $\small C$. 
7. The molar gas constant is  $\small R =$ `r format_dcml(R, size = "small")` $\small\mathit{J}\cdot\mathit{K}^{-1}\mathit{mol}^{-1}$. 





## Numeric input {#numeric-input}

This section begins our detailed discussion of arguments.  

*Scalar input.* &nbsp; Generally used with inline R code. For example, the following R markdown sentence, which includes some math markup and some inline R code,  

````{verbatim}
    The Avogadro constant is $L = $ `r format_sci(L)` $\mathit{mol}^{-1}$. 
````

renders as: The Avogadro constant is $\small L =$ `r format_sci(L, size = "small")` $\small\mathit{mol}^{-1}$.

<br>

*Vector.* &nbsp; A vector of numbers (or a data frame column) is marked up as follows, 

```{r}
# 10. Sample vector
x <- c(2.3333e-5, 3.4444e-4, 5.2222e-2, 6.3333e-1, 8.1111e+1, 9.2222e+2, 2.4444e+4, 3.1111e+5, 4.2222e+6)
format_engr(x)
```

In a table, the output renders as,   

```{r}
#| echo: false
formatdown_options(size = "small")
```

```{r}
DT <- data.table(x, format_engr(x))
knitr::kable(DT,
  align = "r",
  col.names = c("Unformatted", "Engr notation"),
  caption = "Example 10."
)
```

```{r}
#| echo: false
formatdown_options(reset = TRUE)
```

For values with exponents $\small n\in\{-1, 0, 1, 2\}$, the default format is decimal; see [Excluding exponents](#excluding-exponents). 







## Units input

The [units](https://r-quantities.github.io/units/) R package provides measurement units for R vectors, converting vectors of class "numeric" to class "units" [@Pebesma+Mailund+Hiebert:2016:units]. For example

```{r}
# Number
x <- 10320
class(x)

# Convert to units class
units(x) <- "m"
x
class(x)

# Operations are reflected in the values and its units
y <- x^2
y

# Unit conversion is supported
z <- y
z
units(z) <- "ft^2"
z
```

If an input argument to `format_numbers()` (or its convenience functions) is of class "units", formatdown attempts to extract the units character string, format the number in the expected way, and append a units character string to the result. For example, 

```{r}
# 11. Units-class inputs
format_sci(x)
format_sci(y)
format_sci(z)
```

Example 11 renders as, 

- `r format_sci(x, size = "small")`
- `r format_sci(y, size = "small")`
- `r format_sci(z, size = "small")`

More complicated units can be managed. For example the Newtonian gravitational constant could be formatted as follows, where the exponents in the units definition are given in "implicit" form, that is, where $\small\mathrm{m}^3\  \mathrm{kg}^{-1}\mathrm{s}^{-2}$ is represented by `"m3 kg-1 s-2"`. 

```{r}
G <- 6.6743e-11
units(G) <- "m3 kg-1 s-2"
format_sci(G)
```

Applying a similar procedure to several physical constants, we collect the results in a data frame and display in a table to illustrate a variety of units appended to formatted numbers. 

```{r}
#| echo: false
c <- 299792458
units(c) <- "m/s"
format_c <- format_sci(c, size = "small")

h <- 6.62607015e-34
units(h) <- "J/Hz"
format_h <- format_sci(h, size = "small")

mu <- 1.25663706212e-6
units(mu) <- "N A-2"
format_mu <- format_sci(mu, size = "small")

G <- 6.67430e-11
units(G) <- "m3 kg-1 s-2"
format_G <- format_sci(G, size = "small")

ke <- 8.9875517923e+9
units(ke) <- "N m2 C-2"
format_ke <- format_sci(ke, size = "small")

sigma <- 5.67037442e-8
units(sigma) <- "W m-2 K-4"
format_sigma <- format_sci(sigma, size = "small")

symbol <- c(
  "$\\small G$",
  "$\\small c$",
  "$\\small h$",
  "$\\small \\mu_0$",
  "$\\small k_e$",
  "$\\small \\sigma$"
)

quantity <- format_text(
  c(
    "Newtonian gravitational constant",
    "speed of light in a vacuum",
    "Planck constant",
    "vacuum magnetic permeability",
    "Coulomb constant",
    "Stefan-Boltzmann constant"
  ),
  size = "small"
)

formatted_value <- c(
  format_G,
  format_c,
  format_h,
  format_mu,
  format_ke,
  format_sigma
)

DT <- data.table(symbol, quantity, formatted_value)
knitr::kable(DT, caption = "Illustrating a variety of measurement units")
```









## Significant digits {#significant-digits}

Significant digits are applied to the input argument using the base R function `signif()` before additional formatting is applied.  For example, 


```{r}
# 12. Significant digits
format_sci(e, digits = 5)
format_sci(e, digits = 4)
format_sci(e, digits = 3)
```

Example 12 renders as, 

- `r format_sci(e, digits = 5, size = "small")`
- `r format_sci(e, digits = 4, size = "small")`
- `r format_sci(e, digits = 3, size = "small")`






## Formats {#formats}

The `format` argument appears in `format_numbers()` only. The default is "engr". The format is preset in the `format_dcml()`, `format_engr()`, and `format_sci()` convenience functions. 

To compare the effects across many orders of magnitude, we display the same vector in different formats.

```{r}
#| echo: false
formatdown_options(size = "small")
```

```{r}
# 13. Comparing formats
x <- c(2.3333e-5, 3.4444e-4, 5.2222e-2, 6.3333e-1, 8.1111e+1, 9.2222e+2, 2.4444e+4, 3.1111e+5, 4.2222e+6)
dcml <- format_numbers(x, 3, format = "dcml")
sci <- format_numbers(x, 3, format = "sci")
engr <- format_numbers(x, 3, format = "engr")
DT <- data.table(dcml, sci, engr)
knitr::kable(DT,
  align = "r",
  col.names = c("decimal", "scientific", "engineering"),
  caption = "Example 13."
)
```

```{r}
#| echo: false
formatdown_options(reset = TRUE)
```

The values displayed without powers-of-ten notation in the scientific and engineering columns are determined by the `omit_power` argument discussed next.  

## Excluding a range of exponents {#excluding-exponents}

When specifying power-of-ten notation, numbers with exponents lying within the range of the `omit_power` argument are typeset in decimal form. In engineering notation, the exponent is checked for lying within the range before and after the conversion to multiple-of-3 exponents.  

To illustrate, we compare two `omit_power` settings in both scientific and engineering formats. In some columns, we set `omit_power = NULL`, which imposes power-of-ten notation on the entire vector. 

```{r}
#| echo: false
formatdown_options(size = "small")
```

```{r}
# 14. Effects of omit_power
DT <- atmos[3:12, .(pres)]
DT[, sci_all := format_sci(pres, 3, omit_power = NULL)]
DT[, sci_omit := format_sci(pres, 3, omit_power = c(-1, 0))]
DT[, engr_all := format_engr(pres, 3, omit_power = NULL)]
DT[, engr_omit := format_engr(pres, 3, omit_power = c(-1, 0))]
knitr::kable(DT,
  align = "r",
  col.names = c(
    "Unformatted",
    "all scientific",
    "scientific w/ omit",
    "all engineering",
    "engineering w/ omit"
  ),
  caption = "Example 14."
)
```

Comments: 

- In the "scientific with omit" column, the three values $\small (0.184, 1.05, 5.22)$ are in decimal form because their exponents lie within the `omit_power` range $\small \{-1, 0\}$. 
- The same is true in the "engineering with omit" column because these exponents are in the omit range  *before* converting to engineering form. 
- The "engineering with omit" column also displays the values $(\small 22, 80, 287)$ in decimal form because their exponents are in the omit range *after* converting to engineering notation. 

<br>

If a single value is assigned, e.g., `omit_power = 0`, the argument is interpreted as `c(0, 0)`.

```{r}
# 15. Omit power used for a single value of exponent
DT <- atmos[3:12, .(pres)]
DT[, sci_all := format_sci(pres, 3, omit_power = NULL)]
DT[, sci_omit := format_sci(pres, 3, omit_power = 0)]
DT[, engr_all := format_engr(pres, 3, omit_power = NULL)]
DT[, engr_omit := format_engr(pres, 3, omit_power = 0)]
knitr::kable(DT,
  align = "r",
  col.names = c(
    "Unformatted",
    "all scientific",
    "scientific w/ omit",
    "all engineering",
    "engineering w/ omit"
  ),
  caption = "Example 15."
)
```

```{r}
#| echo: false
formatdown_options(reset = TRUE)
```

<br>

Setting `omit_power = c(-Inf, Inf)` yields the same decimal result as `format = "dcml"` and overrides any other `format` setting. For example, 

```{r}
# 16. Different ways of creating a decimal format
(y <- 6.78e-3)

(p <- format_numbers(y, 3, "sci", omit_power = c(-Inf, Inf)))

(q <- format_numbers(y, 3, "dcml"))

(r <- format_dcml(y, 3))

all.equal(p, q)
all.equal(p, r)
```

Example 16 (all cases) renders as, 

- `r format_numbers(y, 3, "dcml", size = "small")`



## Enforcing a specific exponent {#enforcing-exponent}

When values in a table column span only a few orders of magnitude, an audience is sometimes better served by setting the notation to a constant power of ten. For example, here we show numbers in scientific format and compare to columns in which the exponents are set to fixed values. Assigning a value to `set_power` overrides `omit_power` and  `format`. 

```{r}
#| echo: false
formatdown_options(size = "small")
```

```{r}
# 17. set_power argument
DT <- atmos[alt <= 40, .(alt, pres, dens)]
DT[, sci_pres := format_sci(pres, 3, omit_power = c(-1, 2))]
DT[, set_pres := format_sci(pres, 3, omit_power = c(-1, 2), set_power = 3)]
DT[, sci_dens := format_engr(dens, 3, omit_power = c(-1, 2))]
DT[, set_dens := format_engr(dens, 3, omit_power = c(-1, 2), set_power = -2)]
DT[, pres := NULL]
DT[, dens := NULL]
knitr::kable(DT,
  align = "r",
  col.names = c("Altitude (km)", "Pressure (Pa)", "with set_power", "Density (kg/m$^{3}$)", "with set_power"),
  caption = "Example 17."
)
```

```{r}
#| echo: false
formatdown_options(reset = TRUE)
```


## Tables with units

In a typical data table, the numbers in a column have the same physical units and are formatted as a vector. For example, 

```{r}
#| echo: false
formatdown_options(size = "small")
```

```{r}
# Example 18
DT <- air_meas[, .(temp, pres, sp_gas, dens)]

# Examine data
DT[]

# Assign units
units(DT$temp) <- "K"
units(DT$pres) <- "Pa"
units(DT$sp_gas) <- "J kg-1 K-1"
units(DT$dens) <- "kg m-3"

# Make a copy to preserve DT as-is for a subsequent example
x <- copy(DT)

# Format one column at a time
x$temp <- format_dcml(x$temp)
x$pres <- format_engr(x$pres)

# Or format multiple columns in one pass
cols <- c("sp_gas", "dens")
x[, (cols) := lapply(.SD, format_dcml), .SDcols = cols]

knitr::kable(x, align = "r", caption = "Example 18.")
```

```{r}
#| echo: false
formatdown_options(reset = TRUE)
```

<br>

*Placing units in the table header.* &nbsp; An alternate convention is to include the units in the header row only. Here we start with a data frame with unit-class columns. 

```{r}
#| echo: false
formatdown_options(size = "small")
```

```{r}
# Example 19.
y <- copy(DT)
```

First we preserve the columns names and associated measurement units. 

```{r}
# Preserve column names
col_names <- names(y)
col_names

# Preserve units and surround by square brackets for display
unit_list <- sapply(y, units::deparse_unit)
unit_list <- paste0("[", unit_list, "]")
unit_list
```

Drop the units from the numerical columns and format the columns in the usual way. 

```{r}
# Drop units from values
y <- units::drop_units(y)

# Format numerical columns
y$temp <- format_dcml(y$temp)
y$pres <- format_engr(y$pres)
cols <- c("sp_gas", "dens")
y[, (cols) := lapply(.SD, format_dcml), .SDcols = cols]
```

Use `knitr::kable()` and `kableExtra::add_header_above()` to create a two-level header with column names and units. 

```{r}
knitr::kable(y,
  caption = "Example 19.",
  col.names = unit_list,
  align = "r"
) |>
  kableExtra::add_header_above(header = col_names, line = FALSE, align = "r")
```

```{r}
#| echo: false
formatdown_options(reset = TRUE)
```

<br>

*Formatting units in the header.* &nbsp; In the table above, the row of units  is adequate, but the implicit exponents might be confusing to some readers. In this next example, we use formatdown to typeset exponents as superscripts, for example, `[J K-1 kg-1]` becomes $\small\left(\mathrm{J\ K^{-1}\ kg^{-1}}\right)$. 

We start by extracting the first row of the data table, formatting each column using `format_numbers()`, then removing the numerical characters leaving only the math-markup units characters. 

```{r}
# Example 20.
unit_row <- DT[1]
unit_row

# By columns, remove numbers, retain formatted units
for (jj in names(unit_row)) {
  # Select column
  q <- unit_row[, get(jj)]
  # Format the number with its units
  q <- format_numbers(q)
  # Regular expression to identify characters between the $ sign and \\mathrm
  regex <- "(?<=\\$).*?(?=\\\\mathrm)"
  # Remove those characters, leaving the formatted unit string
  unit_row[, jj] <- stringr::str_remove(q, regex)
}
# Convert the data frame row to a vector
unit_vec <- unname(unlist(unit_row[1]))
unit_vec
```

Insert parentheses around the unit expressions. 

```{r}
# Insert parentheses just inside the math delimiters
unit_vec <- stringr::str_replace_all(unit_vec, "\\$", "")
# Add parentheses around units and reinstate the math delimiters
unit_vec <- paste0("$\\left(", unit_vec, "\\right)$")
unit_vec
```

Tidy the column names and use `kableExtra()` to format the table. 

```{r}
# Make the column names presentable
col_names <- names(DT) |>
  stringr::str_replace("temp", "Temperature") |>
  stringr::str_replace("pres", "Pressure") |>
  stringr::str_replace("sp_gas", "Gas constant") |>
  stringr::str_replace("dens", "Density")

# Re-using the formatted data frame "y" from the previous example
# The unit vector is used for table col.name
knitr::kable(y,
  caption = "Example 20.",
  col.names = unit_vec,
  align = "r"
) |>
  # Adjust the font size
  kableExtra::kable_styling(font_size = 13) |>
  # Make the units row a slightly smaller font size
  kableExtra::row_spec(0, font_size = 11) |>
  # Add a header above the units with the variable names
  kableExtra::add_header_above(header = col_names, line = FALSE, align = "r")
```




## Options

Arguments assigned using `formatdown_options()` are described in the [Global settings](global_settings.html) article. 



## References

