---
title: "Format powers of ten"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Format powers of ten}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("formatdown")
library("data.table")

# data.table printout
options(
  datatable.print.nrows = 15,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

The notation we use to represent large and small numbers depends on the context of our communications. In a computer script, for example, we might encode Avogadro's number as `6.0221*10^23`. A computer printout of this number would typically use E-notation, as in `6.0221E+23`.

In professional technical communications, however, computer syntax should be avoided---the asterisk (`*`) and carat (`^`) in `6.0221*10^23` communicate   *instructions to a computer*, not *syntactical mathematics*. And while E-notation (`6.0221E+23`) has currency in some discourse communities, the general convention in technical communications is to format large and small numbers using powers-of-ten notation of the form, 

$$
a \times 10^n, 
$$

The $\times$ multiplication symbol, often avoided in other contexts, is conventional syntax in powers-of-ten notation. Also, the notation has two forms in general use: *scientific* and *engineering* (Chase 2021, 63–67).

- *Scientific.* &nbsp; $n$ is an integer and $a \in Re: 1\leq{|a|}<10$. For example, `r format_power(6.0221E+23, 4, format = "sci")`

- *Engineering.* &nbsp; $n$ is a multiple of 3 and $a \in Re: 1\leq{|a|}<1000$. For example, `r format_power(6.0221E+23, 4)`

*Exceptions.* &nbsp; When exponents are in the neighborhood of zero, for example, $n \in \{-1, 0, 1, 2\}$, decimal notation may be preferred to power of ten notation. Decimal values such as 0.1234, 1.234, 12.34, and 123.4 might be printed as-is. The range of exponents to include in this set is discretionary. 

## Markup

In R Markdown and Quarto Markdown, we use an inline equation markup delimited by `\( ... \)` to create a math expression in the output document (`$ ... $` is also an option). For example, the markup for Avogadro's number is given by, 

        \(N_A = 6.0221\times{10}^{23}\)

which yields \(N_A = 6.0221 \times 10^{23}\) in the document. To *program* the markup, however, we enclose the markup as a character string, that is, 

        "\\(N_A = 6.0221\\times{10}^{23}\\)"

which requires the backward slashes `\` to be "escaped", hence the product symbol is `\\times` and the left and right math delimiters are `\\(` and `\\)`. This is the form of the output produced by `format_power()`. 




## `format_power()`

Given a number, a numerical vector, or a numerical column from a data frame, `format_power()` converts the numbers to character strings of the form, 

        "\\(a\\times{10}^{n}\\)" 
 
where `a` is the coefficient and `n` is the exponent. The user can specify the number of significant digits and scientific or engineering format. Unless otherwise specified, numbers with exponents in the neighborhood of zero are excluded from power of ten notation and are converted to character strings of the form, 

        "\\(a\\)" 

where `a` is the number in decimal notation to the specified number of significant digits. 
 
*Arguments.*

- **x** &nbsp; Numerical vector to be formatted. Can be a scalar, a vector, or a column from a data frame. 

- **digits** &nbsp; Positive nonzero integer to specify the number of significant digits. Default is 3. 

- **format** &nbsp; Possible values are "engr" (engineering notation) and "sci" (scientific notation). Default is "engr". 

- **omit_power** &nbsp; Numeric vector of length two (or NULL). Determines the range of exponents excluded from power-of-ten notation. Default is `c(-1, 2)`. 

- **delim** &nbsp; Character vector of length one or two defining the math markup delimiters. Possible values include `"$"` or `"\\("`, both of which create appropriate left and right delimiters. Custom delimiters can be assigned.

If you are writing your own script to follow along, we use these packages in this vignette:

```{r}
library("formatdown")
library("data.table")
```

*Equivalent usage.*  &nbsp;  The first two arguments do not have to be named if the argument order is maintained. 

```{r}
# Numerical value
avogadro <- 6.0221E+23

# Arguments named
(x <- format_power(x = avogadro, digits = 3))

# Arguments unnamed
y <- format_power(avogadro, 3)

# Implicit use of default argument
z <- format_power(avogadro)

# Demonstrate equivalence
all.equal(x, y)
all.equal(x, z)
```






## Scalar

Use with inline R code.  

```{r}
format_power(avogadro)
```

which, in an Rmd or qmd document, is rendered as `r format_power(avogadro)`. 





## Vector 

```{r}
x <- c(
  1.2222e-6, 2.3333e-5, 3.4444e-4, 4.1111e-3, 5.2222e-2, 6.3333e-1,
  7.4444e+0, 8.1111e+1, 9.2222e+2, 1.3333e+3, 2.4444e+4, 3.1111e+5, 4.2222e+6
)
format_power(x)
```

is rendered as `r format_power(x)[1:3]`, etc.  






## Significant digits

Argument does not have to be named. 

```{r}
format_power(x[1], 3)
format_power(x[1], 4)
```

are rendered as 

- `r format_power(x[1], 3)`
- `r format_power(x[1], 4)`






## Format

Argument, if included, must be named. 

```{r}
format_power(x[3], format = "sci")
format_power(x[3], format = "engr")
```

are rendered as 

- `r format_power(x[3], format = "sci")`
- `r format_power(x[3], format = "engr")`

To compare the effects across many orders of magnitude, we format the example vector twice, placing the results side by side in a data frame for comparison, rendered using `knitr::kable()`, 

```{r}
# Compare two formats
DT <- data.table(
  scientific  = format_power(x, format = "sci"),
  engineering = format_power(x)
)
knitr::kable(DT, align = "rr")
```




## Set or remove exceptions

`omit_power` argument, if included, must be named.   

```{r}
format_power(x[6], omit_power = c(-1, 3))
format_power(x[6], omit_power = c(0, 3))
```

are rendered as 

- `r format_power(x[6], omit_power = c(-1, 3))`
- `r format_power(x[6], omit_power = c(0, 3))`

`omit_power = NULL` removes the exceptions and all numbers are rendered in the selected power-of-ten notation.  

```{r}
# Omit no values from power-of-ten notation
DT <- data.table(
  scientific = format_power(x, format = "sci", omit_power = NULL),
  engineering = format_power(x, omit_power = NULL)
)
knitr::kable(DT, align = "rr")
```





## Delimiters

`delim` argument, if included, must be named. 

```{r}
format_power(x[1], delim = "\\(")
format_power(x[1], delim = c("\\(", "\\)"))
```

are both rendered as 

- `r format_power(x[1], delim = "\\(")`

In most output documents, the `$ ... $` delimiters yield the same results. 

```{r}
format_power(x[1], delim = "$")
format_power(x[1], delim = c("$", "$"))
```

are both rendered as 

- `r format_power(x[1], delim = "$")`

I encountered one instance in which the `$ ... $` delimiters did not produce the desired output---using `kableExtra::kbl()` to create a table in a `.qmd` output document. Switching the delimiters to `\\( ... \\)` fixed the problem, per the recommendation of the MathJax consortium  (Cervone, 2018). 




## Data frame

Using `density`, a data frame included with formatdown that contains columns of class Date, character, factor, numeric, and integer.

```{r}
density
```

which is rendered un-formatted as

```{r}
#| echo: false
knitr::kable(density, align = "ccrrr")
```

Treating a column as a vector, 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Format as a vector
format_power(DT$p_Pa, digits = 4)
```

is rendered as `r format_power(DT$p_Pa, digits = 4)`. 

Treating a column within a data frame, 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Format one column, retain all columns
DT$p_Pa <- format_power(DT$p_Pa, digits = 4)
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "r")
```

Using `lapply()` to select and treat multiple columns from a data frame, 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Identify columns to format
cols_we_want <- c("T_K", "p_Pa", "density")

# Select and format.
DT <- DT[, lapply(.SD, function(x) format_power(x, 4)), .SDcols = cols_we_want]
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "r")
```

Repeat, but retain all columns,  

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Identify columns to format
cols_we_want <- c("T_K", "p_Pa", "density")

# Format selected columns, retain all columns
DT <- DT[, (cols_we_want) := lapply(.SD, function(x) format_power(x, 4)), .SDcols = cols_we_want]
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "ccrrr")
```

If different columns should be reported with different significant digits, treat the columns separately. 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Format selected columns, retain all columns with signif digits = 5
cols_we_want <- c("T_K", "density")
DT <- DT[, (cols_we_want) := lapply(.SD, function(x) format_power(x, 5)), .SDcols = cols_we_want]

# Individually format one column with signif digits = 4
DT$p_Pa <- format_power(DT$p_Pa, 4)
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "ccrrr")
```








## Non-numeric input

Applying the function to non-numeric objects produces errors. For example, "Date" class. 

```{r}
#| error: true
x <- density$date
format_power(x)
```

Error on "character" class, 

```{r}
#| error: true
x <- density$trial
format_power(x)
```

Error on "factor" class, 

```{r}
#| error: true
x <- density$humidity
format_power(x)
```





## References

Davide Cervone (2018) MathJax: TeX and LaTeX math delimiters,   https://docs.mathjax.org/en/v2.7-latest/tex.html#tex-and-latex-math-delimiters. 

Morgan Chase (2021) *Technical Mathematics,*  https://openoregon.pressbooks.pub/techmath/chapter/module-11-scientific-notation/.

