---
title: "Format powers of ten"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Format powers of ten}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("formatdown")
library("data.table")

# data.table printout
options(
  datatable.print.nrows = 15,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

The notation we use to represent large and small numbers depends on the communications context. In a computer script, for example, we might encode Avogadro's number as `6.0221*10^23`. A computer printout of this number would typically use E-notation, as in `6.0221E+23`.

In professional technical communications, however, computer syntax should be avoided---the asterisk (`*`) and carat (`^`) in `6.0221*10^23` *encode  instructions* to a computer rather than *communicating mathematics* to a  reader. And while E-notation (`6.0221E+23`) has currency in some discourse communities, the general convention in technical communications is to format large and small numbers using powers-of-ten notation of the form, 

$$
a \times 10^n, 
$$

The $\times$ multiplication symbol, often avoided in other contexts, is conventional syntax in powers-of-ten notation [@Chase:2021, 63--67]. Moreover, power-of-tens notation has two formulations: *scientific notation* and *engineering notation*.

- *Scientific.* &nbsp; $n$ is an integer and $a \in \mathbb{R}: 1\leq{|a|}<10$. For example, `r format_power(6.0221E+23, 4, format = "sci")`

- *Engineering.* &nbsp; $n$ is a multiple of 3 and $a \in \mathbb{R}: 1\leq{|a|}<1000$. For example, `r format_power(6.0221E+23, 4)`

When exponents are in the neighborhood of zero, for example, $n \in \{-1, 0, 1, 2\}$, decimal notation may be preferred.  Thus decimal values 0.1234, 1.234, 12.34, and 123.4 will often be printed as-is. The range of this set is usually adjusted to suit one's audience and rhetorical goals. 

## Markup

In R Markdown and Quarto Markdown, we use an inline equation markup delimited by `$...$` to create a math expression in the output document. For example, the markup for Avogadro's number is given by, 

        $N_A = 6.0221\times{10}^{23}$

which yields $N_A = 6.0221 \times 10^{23}$ in the document. To *program* the markup, however, we enclose the markup as a character string, that is, 

        "$N_A = 6.0221\\times{10}^{23}$"

which requires the backward slash `\` to be "escaped", hence the product symbol is denoted `\\times`. This is the form of the output produced by `format_power()`. 




## `format_power()`

Given a number, a numerical vector, or a numerical column from a data frame, `format_power()` converts the numbers to character strings of the form, 

        "$a\\times{10}^{n}$"
 
where `a` is the coefficient and `n` is the exponent. The user can specify the number of significant digits and scientific or engineering format. Over the range defined for decimal notation, the numbers are converted to character strings of the form 

        "$a$"

where `a` is the number in decimal notation to the specified number of significant digits. 
 
*Arguments.*

- **x** &nbsp; Numerical vector to be formatted. Can be a scalar, a vector, or a column from a data frame. 

- **digits** &nbsp; Positive nonzero integer to specify the number of significant digits. Default is 3. 

- **format** &nbsp; Possible values are "engr" (engineering notation) and "sci" (scientific notation). Default is "engr". 

- **omit_power** &nbsp; Numeric vector of length two, default `c(-1, 2)`. The range of exponents over which power of ten notation is omitted and numbers are formatted in decimal form. 

If you are writing your own script to follow along, we use these packages in this vignette:

```{r}
library("formatdown")
library("data.table")
```

*Equivalent usage.*  &nbsp;  The first two arguments do not have to be named if the argument order is maintained. 

```{r}
# Numerical value
avogadro <- 6.0221E+23

# Arguments named
(x <- format_power(x = avogadro, digits = 3))

# Arguments unnamed
y <- format_power(avogadro, 3)

# Implicit use of default argument
z <- format_power(avogadro)

# Demonstrate equivalence
all.equal(x, y)
all.equal(x, z)
```






## Examples

*Scalar.* &nbsp;  Use with inline R code.  

```{r}
format_power(avogadro)
```

which, in an Rmd or qmd document, is rendered as `r format_power(avogadro)`. 

<br>*Vector.* &nbsp;  

```{r}
x <- c(
  1.2222e-6, 2.3333e-5, 3.4444e-4, 4.1111e-3, 5.2222e-2, 6.3333e-1,
  7.4444e+0, 8.1111e+1, 9.2222e+2, 1.3333e+3, 2.4444e+4, 3.1111e+5, 4.2222e+6
)
format_power(x)
```

is rendered as `r format_power(x)[1:3]`, etc.  

<br>*Significant digits.* &nbsp; Argument does not have to be named. 

```{r}
format_power(x[1], 3)
format_power(x[1], 4)
```

are rendered as 

- `r format_power(x[1], 3)`
- `r format_power(x[1], 4)`

<br>*Format.* &nbsp; Argument, if included, must be named. 

```{r}
format_power(x[3], format = "sci")
format_power(x[3], format = "engr")
```

are rendered as 

- `r format_power(x[3], format = "sci")`
- `r format_power(x[3], format = "engr")`


<br>*Other_than* &nbsp; Argument, if included, must be named.   

```{r}
format_power(x[6], omit_power = c(-1, 3))
format_power(x[6], omit_power = c(0, 3))
```

are rendered as 

- `r format_power(x[6], omit_power = c(-1, 3))`
- `r format_power(x[6], omit_power = c(0, 3))`


```{r}
format_power(x[6], omit_power = c(1, 1))
```





To compare the effects across many orders of magnitude, we treat the example vector but convert it to a one-column data frame for display, rendered using using `knitr::kable()`, 

```{r}
# Format using defaults and convert to a data frame
DT <- data.table(
  scientific = format_power(x, format = "sci"),
  engineering = format_power(x)
)
knitr::kable(DT, align = "rr")
```

Repeat but with all numbers of powers-of-ten notation. 

```{r}
# Omit all decimal notations
DT <- data.table(
  scientific = format_power(x, format = "sci", omit_power = NULL),
  engineering = format_power(x, omit_power = NULL)
)
knitr::kable(DT, align = "rr")
```







<br>*Data frame.* &nbsp; Using `density`, a data frame included with formatdown that contains columns of class Date, character, factor, numeric, and integer.

```{r}
density
```

which is rendered un-formatted as

```{r}
#| echo: false
knitr::kable(density, align = "ccrrr")
```

Treating a column as a vector, 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Format as a vector
format_power(DT$p_Pa, digits = 4)
```

is rendered as `r format_power(DT$p_Pa, digits = 4)`. 

Treating a column within a data frame, 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Format one column, retain all columns
DT$p_Pa <- format_power(DT$p_Pa, digits = 4)
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "r")
```

Using `lapply()` to select and treat multiple columns from a data frame, 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Identify columns to format
cols_we_want <- c("T_K", "p_Pa", "density")

# Select and format.
DT <- DT[, lapply(.SD, function(x) format_power(x, 4)), .SDcols = cols_we_want]
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "r")
```

Repeat, but retain all columns,  

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Identify columns to format
cols_we_want <- c("T_K", "p_Pa", "density")

# Format selected columns, retain all columns
DT <- DT[, (cols_we_want) := lapply(.SD, function(x) format_power(x, 4)), .SDcols = cols_we_want]
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "ccrrr")
```

If different columns should be reported with different significant digits, treat the columns separately. 

```{r}
# Copy to avoid "by reference" changes to density
DT <- copy(density)

# Format selected columns, retain all columns with signif digits = 5
cols_we_want <- c("T_K", "density")
DT <- DT[, (cols_we_want) := lapply(.SD, function(x) format_power(x, 5)), .SDcols = cols_we_want]

# Individually format one column with signif digits = 4
DT$p_Pa <- format_power(DT$p_Pa, 4)
DT[]
```

is rendered as

```{r}
#| echo: false
knitr::kable(DT, align = "ccrrr")
```

*Non-numeric input.* &nbsp; Applying the function to non-numeric objects produces errors. For example, "Date" class. 

```{r}
#| error: true
x <- density$date
format_power(x)
```

Error on "character" class, 

```{r}
#| error: true
x <- density$trial
format_power(x)
```

Error on "factor" class, 

```{r}
#| error: true
x <- density$humidity
format_power(x)
```





## References

<div id="refs"></div>




